---
title: "Lexis diagrams: cancer incidence and mortality"
subtitle: "`r params$country`"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    code_folding: hide
    self_contained: true
params:
  country: "Germany"
  site: "all"
  sex: 0
  age_min: 20
  age_max: 84
  delta: 5
  panel_min_years: 20
---

<!-- Render: Rscript code/09_render_lexis.R Germany -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6.5, dpi = 150)
library(data.table); library(ggplot2); library(here)
source(here("code", "functions_asr.R"))
library(matisseR)   # MATISSE port (remotes::install_github("filhoalm/matisseR"))

file_inc  <- here("data", "ECIS", "ecis_historical_data.csv")
file_mort <- here("data", "who_mortality", "data_full_history.rds")
file_pop  <- here("data", "who_mortality", "population_full_history.rds")
file_ctry <- here("data", "dict", "g2022_country_info.csv")
file_site <- here("data", "dict", "site_map.csv")
dir_out   <- here("output", "lexis")
dir.create(dir_out, showWarnings = FALSE, recursive = TRUE)

st <- fread(file_site, colClasses = "character", na.strings = NULL)[site == params$site]
stopifnot(nrow(st) == 1, params$age_max < 85, params$delta %% 5 == 0)
dt_ent <- st[, .(entity = strsplit(ecis_entity, ";")[[1]]), by = site]
codes  <- as.integer(strsplit(st$who_codes, ";")[[1]])
sexes  <- as.integer(strsplit(st$sexes, ";")[[1]])
stopifnot(params$sex %in% c(if (length(sexes) == 2) 0L, sexes))

age_groups <- (params$age_min %/% 5 + 1):(params$age_max %/% 5 + 1)
sex_lab <- c("0" = "Both sexes", "1" = "Males", "2" = "Females")[[as.character(params$sex)]]
tag <- tolower(gsub("\\W+", "_", paste(params$country, params$site, c("MF", "M", "F")[params$sex + 1])))
fullname <- sprintf("%s, %s, %s", params$country, st$site_label, sex_lab)
```

## Lexis diagrams in MATISSE and here

MATISSE (Rosenberg; methods in Miranda Filho & Rosenberg, *Front Oncol* 2024)
represents a Lexis diagram as a `rates` object: an age x period array of square
cells holding events and person-years. The functions in
the matisseR package reproduce the MATISSE methods used in its Lexis diagram
tutorial (`MATISSE_2_The_Lexis_Diagram.mlx`):

| MATISSE | R (matisseR) | What it does |
|---|---|---|
| `rates(...)`, `csv2rates` | `rates_make()` | Events and Offset matrices, age and period cutpoints |
| `chunk(r, 'AgeBlock', k, 'PerBlock', k)` | `rates_chunk()` | Aggregate into k-year blocks; leftover early years dropped (LTF) |
| `fill(r)` | `rates_fill()` | Zero events replaced by the adaptive small value (>= 0.5) |
| `uni(r)` | `rates_uni()` | Summary statistics |
| `type(r, 'comp', 'lr')` | `rates_long()` | Cell rates (log scale in the plots) |
| `lexis(r, 'pcolor')` | `lexis_plot()` | Age x period beside age x cohort, same aspect ratio |
| `rates2csv` | `rates2csv()` | NCI APC Web Tool / MATISSE CSV file |

MATISSE needs square cells. ECIS and WHO report 5-year age groups, so calendar
years are chunked into `r params$delta`-year periods, giving
`r params$delta` x `r params$delta` cells. Cohorts are labelled by their
midpoint (period midpoint minus age midpoint) and span `r 2 * params$delta` years.

## Data

**`r fullname`**, ages `r params$age_min`-`r params$age_max`.

- **Incidence**: ECIS historical data, *`r gsub(";", " + ", st$ecis_entity)`*, summed over the
  registries reporting every year of the common panel with unchanged coverage
  (a registry whose covered population changes by more than 20% in a year, such as
  North Rhine-Westphalia in 2010, is treated as a new unit from that year).
- **Mortality**: WHO Mortality Database, national,
  `r if (length(codes)) paste0("ICD-10 ", st$icd10) else "not available for this site (no separate WHO code)"`.

```{r data}
inc   <- ecis_country_incidence(file_inc, params$country, dt_ent, params$panel_min_years)
R_inc <- rates_fill(rates_chunk(rates_make(inc$data[params$sex == 0 | sex == params$sex], age_groups,
                                           paste(fullname, "incidence"), "Cases"),
                                age_block = params$delta / 5, per_block = params$delta))

R_mort <- NULL
if (length(codes)) {
  dt_mort <- who_mortality(file_mort, file_pop, country_m49(file_ctry, params$country),
                           data.table(site = params$site, cancer_code = codes))
  R_mort  <- rates_fill(rates_chunk(rates_make(dt_mort[params$sex == 0 | sex == params$sex], age_groups,
                                               paste(fullname, "mortality"), "Deaths"),
                                    age_block = params$delta / 5, per_block = params$delta))
}

knitr::kable(rbindlist(list(rates_uni(R_inc), if (!is.null(R_mort)) rates_uni(R_mort))),
             format.args = list(big.mark = ","), caption = "Rates objects (cf. MATISSE uni).")
```

Incidence registry panel: `r inc$panel$sel$n_registry` registries,
`r inc$panel$sel$y0`-`r inc$panel$sel$y1` (`r paste(inc$panel$sel$registries[[1]], collapse = "; ")`).

## Incidence

```{r lexis-inc, fig.cap = "Incidence: log rates on the Lexis diagram (MATISSE lexis method)."}
lexis_plot(R_inc, cols = c("#cde2fb", "#0d366b"), legend = "Incidence per 100,000")
```

## Mortality

```{r lexis-mort, eval = !is.null(R_mort), fig.cap = "Mortality: log rates on the Lexis diagram (MATISSE lexis method)."}
lexis_plot(R_mort, cols = c("#fde4d6", "#8f2d08"), legend = "Mortality per 100,000")
```

Reading the diagrams: in the left panel each column is an age group and each row
a calendar period; in the right panel the same cells are re-stacked by birth
cohort, so a horizontal row follows one cohort as it ages. A cohort effect shows
as colour changing along rows on the right, a period effect along rows on the left.

## Outputs for MATISSE and the NCI APC Web Tool

```{r out}
f_inc <- file.path(dir_out, sprintf("%s_incidence_%dx%d.csv", tag, params$delta, params$delta))
rates2csv(R_inc, f_inc, R_inc$fullname, sprintf("ECIS incidence, registry panel %d-%d", inc$panel$sel$y0, inc$panel$sel$y1))
files <- f_inc
if (!is.null(R_mort)) {
  f_mort <- file.path(dir_out, sprintf("%s_mortality_%dx%d.csv", tag, params$delta, params$delta))
  rates2csv(R_mort, f_mort, R_mort$fullname, "WHO Mortality Database, national")
  files <- c(files, f_mort)
}
knitr::kable(data.table(file = basename(files)))
```

Each file loads into MATISSE with `R = csv2rates('<file>.csv')` (header lines
Title, Description, Start Year, Start Age, Interval, then paired events and
person-years columns by period, one row per age group), and into the NCI APC Web
Tool (<https://analysistools.cancer.gov/apc/>).

## Provenance

```{r prov}
dt_files <- data.table(path = c(file_inc, file_mort, file_pop, file_site))
dt_files[, `:=`(file = basename(path), size_mb = round(file.size(path) / 2^20, 1),
                md5 = unname(tools::md5sum(path)), modified = format(file.mtime(path)))]
knitr::kable(dt_files[, .(file, size_mb, md5, modified)])
```

MATISSE: Rosenberg PS, MATLAB toolbox for vital rates surveillance; methods in
Miranda Filho A, Rosenberg PS, *Front Oncol* 2024, doi:10.3389/fonc.2023.1332429.

```{r session}
sessionInfo()
```
